
---
title: "CHOP metadata checker"
author: "PennCHOP Microbiome Program"
output:
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '3'
geometry: margin=3cm
---

<script>
$(document).ready(function(){
    $('[data-toggle="popover"]').popover(); 
});
</script>

```{r setup, echo=FALSE}
### ================
###   knitr setup
### ================
library(knitr)
opts_chunk$set(
  tidy=FALSE,
  cache=TRUE,
  echo=FALSE,
  warning=FALSE,
  message=FALSE,
  dpi=100,
  fig.width=7,
  fig.height=8,
  fig.align = "center"
  )
options(knitr.table.format = "html")
```

```{r libraries, message=FALSE, warning=FALSE}

library(magrittr)
library(tidyverse)
library(kableExtra)

```

# Introduction

This [website](/) is used to validate your metadata spreadsheet used in bioinformatics analyses for the PennCHOP Microbiome Center. This checker will parse your spreadsheet to ensure it fulfills the specifications listed below and highlight and display error messages for cells that do not meet these specifications. All that is required is a tab or comma separatated file (tsv/csv) to be uploaded.
\
\

# Specifications

### Mandatory Columns
The contents within these columns cannot be empty
```{r}
manda_cols <- as.data.frame(matrix(c(
  "SampleID",
  "investigator",
  "project_name",
  "sample_type",
  "tube_barcode",
  "box_id",
  "box_position",
  "study_group",
  "date_collected"
  ), ncol = 3))

manda_cols %>%
  select_all(~gsub("_", "\\\\_", .)) %>% ##need to escape the underscores
  mutate_if(function(x) is.character(x) | is.factor(x), ~gsub("_", "\\\\_", .)) %>%
  kable("html", longtable = F, digits=2, booktabs=T, escape=F, col.names = NULL) %>%
  kable_styling(latex_options = c("scale_down", "repeat_header"))

```

### Suggested Columns

```{r}
sugg_cols <- as.data.frame(matrix(c(
  "subject_id",
  "host_species",
  "study_day",
  "current_antibiotics",
  "recent_antibiotics",
  "cage_id",
  "mouse_strain",
  ""), ncol = 4))

sugg_cols %>%
  select_all(~gsub("_", "\\\\_", .)) %>% ##need to escape the underscores
  mutate_if(function(x) is.character(x) | is.factor(x), ~gsub("_", "\\\\_", .)) %>%
  kable("html", longtable = F, digits=2, booktabs=T, escape=F, col.names = NULL) %>%
  kable_styling(latex_options = c("scale_down", "repeat_header"))

```

### Formatting

##### General cell formatting
All cells can only contain numbers, characters, spaces and the following special characters: ._+-/<>=|,()[]
\
\

##### Special cell formatting
[SampleID] must begin with a letter and only contain numbers, characters, periods, and underscores (no other special characters)\
[subject_id] can only contain numbers, letters, periods, dashes, and underscores\
[date_collected] can only be in format yyyy-mm-dd\
[time_collected] can only be in format hh:mm:ss\
[barcode] can only contain nucleotide symbols\
[forward_barcode_location] and [reverse_barcode_location] can only contain 96-well plate IDs
\
\

##### Column name formatting
The names of the columns may only contain numbers, letters, and underscores
\
\

##### Unique combinations
Contents between these column pairs must be unique together:
```{r}
uniq_combo <- as.data.frame(matrix(c("box_id", "box_position",
                                     "reverse_barcode_plate", "reverse_barcode_location",
                                     "forward_barcode_plate", "forward_barcode_location"), ncol = 2, byrow = TRUE))

uniq_combo %>%
  select_all(~gsub("_", "\\\\_", .)) %>% ##need to escape the underscores
  mutate_if(function(x) is.character(x) | is.factor(x), ~gsub("_", "\\\\_", .)) %>%
  kable("html", longtable = F, digits=2, booktabs=T, escape=F, col.names = NULL) %>%
  kable_styling(latex_options = c("scale_down", "repeat_header"))

```

### Pre-selected options
If you do not see an option listed here for your sample, please contact us

##### [host_species] options
```{r}
host_df <- as.data.frame(matrix(c(
  "Dog",
  "Fruit fly",
  "Human",
  "Mouse",
  "Naked mole rat",
  "Pig",
  "Pigeon",
  "Rabbit",
  "Rat",
  "Rhesus macaque",
  "None",
  ""), ncol = 4))

host_df %>%
  select_all(~gsub("_", "\\\\_", .)) %>% ##need to escape the underscores
  mutate_if(function(x) is.character(x) | is.factor(x), ~gsub("_", "\\\\_", .)) %>%
  kable("html", longtable = F, digits=2, booktabs=T, escape=F, col.names = NULL) %>%
  kable_styling(latex_options = c("scale_down", "repeat_header"))

```

##### [sample_type] options

```{r}
sample_df <- as.data.frame(matrix(c(
  "Amnoitic fluid",
  "BAL",
  "Bedding",
  "Biofilm",
  "Bioreactor",
  "Blank swab",
  "Blood",
  "Breast milk",
  "Buffer",
  "Cecum",
  "Cell lysate",
  "Cervical swab",
  "Cheek swab",
  "Crop",
  "Dental plaque",
  "Duodenum",
  "Dust",
  "Elution buffer",
  "Empty well",
  "Endometrial swab",
  "Environmental control",
  "Esophageal biopsy",
  "Esophagus",
  "Feces",
  "Feed",
  "Fistula",
  "Fistula swab",
  "Fly food",
  "Fruit fly",
  "Ileostomy fluid",
  "Ileum",
  "Kveim reagent",
  "Lab water",
  "Macular Retina",
  "Meconium",
  "Medium",
  "Microbial culture",
  "Mock DNA",
  "Mouse chow",
  "Nasal swab",
  "Nasopharyngeal swab",
  "Oral swab",
  "Oral wash",
  "Oropharyngeal swab",
  "Ostomy fluid",
  "Pancreatic fluid",
  "PCR water",
  "Peripheral retina",
  "Placenta",
  "Plasma",
  "Rectal biopsy",
  "Rectal swab",
  "Saline",
  "Saliva",
  "Sediment",
  "Serum",
  "Skin swab",
  "Small intestine",
  "Soil",
  "Sputum",
  "Surface swab",
  "Tongue swab",
  "Tonsil",
  "Tracheal aspirate",
  "Tracheal control",
  "Urethral swab",
  "Urine",
  "Water",
  "Weighing paper",
  "Whole gut",
  "Large intestine mucosa",
  "Large intestine lumen"
  ), ncol = 6))

sample_df %>%
  select_all(~gsub("_", "\\\\_", .)) %>% ##need to escape the underscores
  mutate_if(function(x) is.character(x) | is.factor(x), ~gsub("_", "\\\\_", .)) %>%
  kable("html", longtable = F, digits=2, booktabs=T, escape=F, col.names = NULL) %>%
  kable_styling(latex_options = c("scale_down", "repeat_header"))

```

\

# Example of misformatted sheet
Misformatted cells are highlighted in red

```{r}

df_wrong <- data.frame(`sample_id` = c('Id1', 'Id2', 'Id2'),
                  `host_species` = c('Mouse', 'Turtle', 'Mouse'),
                  `study_group` = c('Ctrl', 'Abx', '@bx'),
                 `box_id` = c('Box1', 'Box1', 'Box1'),
                 `box_position` = c('A1', 'A1', 'A2'))

df_wrong %>%
  select_all(~gsub("_", "\\\\_", .)) %>% ##need to escape the underscores
  mutate_if(function(x) is.character(x) | is.factor(x), ~gsub("_", "\\\\_", .)) %>%
  kable("html", longtable = F, digits=2, booktabs=T, escape=F) %>%
  kable_styling(latex_options = c("scale_down", "repeat_header")) %>%
  column_spec(1,
              background = ifelse(duplicated(df_wrong$sample_id)|duplicated(df_wrong$sample_id, fromLast = TRUE), "#f5c6ca", "#FFFFFF"),
              popover = ifelse(duplicated(df_wrong$sample_id)|duplicated(df_wrong$sample_id, fromLast = TRUE), "Please use unique values", "")) %>%
  column_spec(2,
              background = ifelse(!df_wrong$host_species %in% c("Human", "Mouse", "Rat", "NA"), "#f5c6ca", "#FFFFFF"),
              popover = ifelse(!df_wrong$host_species %in% c("Human", "Mouse", "Rat", "NA"), "Please use only pre-selected host_species", "")) %>%
  column_spec(3,
              background = ifelse(!grepl("^[0-9A-Za-z._+-/<>=|,() ]+$", df_wrong$study_group), "#f5c6ca", "#FFFFFF"),
              popover = ifelse(!grepl("^[0-9A-Za-z._+-/<>=|,() ]+$", df_wrong$study_group), "Please use the correct formatting specified above", "")) %>%
  column_spec(4:5,
              background = ifelse(duplicated(paste(df_wrong$box_id, df_wrong$box_position))|duplicated(paste(df_wrong$box_id, df_wrong$box_position), fromLast = TRUE), "#f5c6ca", "#FFFFFF"),
              popover = ifelse(duplicated(paste(df_wrong$box_id, df_wrong$box_position))|duplicated(paste(df_wrong$box_id, df_wrong$box_position), fromLast = TRUE), "Please use unique values", ""))
  


```
\

# Example of correctly formatted sheet

```{r}

df_right <- data.frame(`sample_id` = c('Id1', 'Id2', 'Id3'),
                       `host_species` = c('Mouse', 'Mouse', 'Mouse'),
                       `study_group` = c('Ctrl', 'Abx', 'Abx'),
                       `box_id` = c('Box1', 'Box1', 'Box1'),
                       `box_position` = c('A1', 'A2', 'A3'))

df_right %>%
  select_all(~gsub("_", "\\\\_", .)) %>%
  mutate_if(function(x) is.character(x) | is.factor(x), ~gsub("_", "\\\\_", .)) %>%
  kable("html", longtable = F, digits=2, booktabs=T, escape=F) %>%
  kable_styling(latex_options = c("scale_down", "repeat_header"))

```

#### [Return to checker](/)
